Hazelnut: A Bidirectionally Typed \\ Structure Editor Calculus

### 2016-07-05 example.tex
tweaks to sec 2 and 3

Figure~\ref{fig:first-example} shows an edit sequence that constructs the increment function, of type $\tarr{\tnum}{\tnum}$, starting from an empty hole via a sequence of {actions}. We will discuss the syntactic forms and the referenced rules in Sec. \ref{sec:hazel}. For now, let us develop only the high-level intuitions.

@@

The edit state in Hazelnut is a {Z-expression}, $\zexp$. Every Z-expression has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the {cursor}. The selected H-expression or H-type is typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively. For example, on Line 1 of Figure \ref{fig:first-example}, the empty expression hole, $\hhole{}$, is under the cursor.

@@

\begin{figure}[t!]
\begin{center}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} & 
%\textbf{H-Expression} & 
% \textbf{Type} & 
\textbf{Next Action} & \textbf{Rule}
\\
\hline
1 &
\zwsel{\hhole{}} & 
% \hhole{} &
% \tehole 
% &
\aConstruct{\flam{x}} & 
\text{(\ref{r:conelamhole})}
\\ 2 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} & 
% \hlam{x}{\hhole{}} : \tarr{\hhole{}}{\hhole{}} &
% \tarr{\tehole}{\tehole} &
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 3 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\hhole{}} &
% \tarr{\tnum}{\tehole} & 
\aMove{\dNext{}} & 
\text{({\ref{rule:move-nextSib-arr}})}
\\ 4 &
\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
&
% \text{\textquotedbl}&
% \tarr{\tnum}{\tehole} & 
\aConstruct{\fnum{}} & 
\text{(\ref{r:contnum})}
\\ 5 &
\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} & 
% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\tnum{}} &
% \tarr{\tnum}{\tnum} &
\aMove{\dParent{}} & 
\text{(\ref{rule:move-parent-arr-right})}
\\ 6 &
\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
&
% \text{\textquotedbl}&
% \tarr{\tnum}{\tnum} & 
\aMove{\dParent{}} & 
\text{(\ref{rule:move-parent-asc-right})}
\\ 7 & 
\zwsel{\hlam{x}{\hhole{}} : \tarr{\tnum}{\tnum}} & 
\aMove{\dChild} & 
\text{(\ref{r:movefirstchild})}
\\ 8 &
% &
\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} & 
% \tarr{\tnum}{\tnum} &
% &
\aMove{\dChild{}} & 
\text{(\ref{r:movefirstchild-lam})}
\\ 9 &
% &
\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
% \tarr{\tnum}{\tnum} & 
\aConstruct{\fvar{x}} & 
\text{(\ref{r:conevar})}
\\ 10 &
% \hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
% &
\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
% \tarr{\tnum}{\tnum} & 
{\aConstruct{\fplus}}
&
\text{(\ref{rule:construct-plus-compat})}
\\ 11 & 
\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} & 
\aConstruct{\fnumlit{1}} & 
\text{(\ref{r:conenumnum})}
\\ 12 & 
\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}} : \tarr{\tnum}{\tnum} & 
\textrm{---} & 
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Constructing an increment function in Hazelnut. The formal syntax and the rules referenced in the final column are described in Section \ref{sec:hazel}.}
\label{fig:first-example}
\end{figure}

The actions on Lines 5-8 move the cursor to the function body. For simplicity, Hazelnut defines only a small collection of primitive movement actions. In practice, an editor would also support compound movement actions, e.g. an action that moves the cursor directly to the next remaining hole, defined in terms of these primitive movement actions.

Lines 9-11 complete the function body by first constructing the variable $x$, then  constructing the plus form, and finally constructing the number $\hnum{1}$. Notice that we did not need to construct the function body in an ``outside-in'' manner, i.e. we constructed $x$ before constructing the outer plus form. The transient function body, $x$, was coincidentally of type $\tnum$, so Hazelnut needed to make no special allowances in this case.

@@

\begin{figure}[t!]
\begin{center}
\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{53px}}
% \vspace{-6px}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
% \\
\hline
\# & \textbf{Z-Expression} & 
%\textbf{H-Expression} & 
% \textbf{Type} & 
\textbf{Next Action} & \textbf{Rule}
\\
\hline
13 &
\zwsel{\hhole{}} & 
% \hhole{} &
% \tehole 
% &
\aConstruct{\fvar{incr}} \hphantom{~\,\,~~}& 
\text{(\ref{r:conevar})}
\\ 14 & 
\zwsel{incr} & 
\aConstruct{\fap} & 
\text{(\ref{r:coneapfn})}
\\ 15 & 
incr(\zwsel{\hhole{}}) & 
\aConstruct{\fvar{incr}} & 
\text{(\ref{r:conevar2})}
\\ 16 & 
incr(\hhole{\zwsel{incr}}) & 
\aConstruct{\fap} & 
\text{(\ref{r:coneapfn})}
\\ 17 & 
incr(\hhole{incr(\zwsel{\hhole{}})}) \hphantom{~~~~} & 
\aConstruct{\fnumlit{3}} & 
\text{(\ref{r:conenumnum})}
\\ 18 & 
incr(\hhole{incr(\zwsel{\hnum{3}})}) & 
\aMove{\dParent}& 
\text{(\ref{r:moveparent-ap2})}
\\ 19 & 
incr(\hhole{\zwsel{incr(\hnum{3})}}) & 
\aMove{\dParent} & 
\text{(\ref{r:moveparent-hole})}
\\ 20 &
incr(\zwsel{\hhole{incr(\hnum{3})}})& 
\aFinish & 
\text{(\ref{r:finishana})}
\\ 21 & 
incr(\zwsel{incr(\hnum{3})}) & 
\textrm{---} & 
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Appyling the increment function.}
\label{fig:second-example}
\end{figure}

Figure \ref{fig:second-example} shows an edit sequence that constructs the expression $incr(incr(\hnum{3}))$, where $incr$ is assumed bound to the increment function from Figure \ref{fig:first-example} (we omit these steps for concision.)

Line 13 constructs a variable identifying the function we wish to apply, i.e. $incr$. Line 14 then constructs the application form with $incr$ in function position, leaving the cursor on a hole in the argument position.

We need to apply $incr$ again, so we perform the same action on Line 15 as we did on Line 13, i.e. $\aConstruct{\fvar{incr}}$. Na\"ively, however, this would result in the following edit state: 
\[
incr(\zwsel{incr})
\]
This edit state is ill-typed, i.e. the argument of $incr$ must be of type $\tnum$ but is here of type $\tarr{\tnum}{\tnum}$. To maintain our edit invariant, this edit state cannot be allowed to arise.

The programmer could alternatively have performed the $\aConstruct{\fap}$ action on Line 15. This would result in the following edit state, which is well-typed according to the static semantics we will define in the next section:
\[
incr(\hhole{}(\zwsel{\hhole{}}))
\]
The problem is that the programmer is not able to identify the intended function before constructing the outer application form. This is qualitatively unnatural.

Hazelnut supports a more flexible strategy: rather than disallowing the $\aConstruct{\fvar{incr}}$ action on Line 15, it puts $incr$ inside a hole:
\[
incr(\hhole{\zwsel{incr}})
\]
As we will detail, this defers the necessary type compatibility check, exactly as the empty hole does. One way to think about non-empty holes is as a formalization of the heuristic concept of the ``red underline'' that a text or syntactic structure editor displays where a type incompatibility is detected. The difference is that discovering a type incompatibility does not leave the entire program formally meaningless.

The expression inside a non-empty hole must itself be well-typed, so we can continue to edit it. Lines 16-17 proceed to apply the inner mention of $incr$ to a number literal, $\hnum{3}$. Finally, Lines 18-19 move the cursor to the non-empty hole and Line 20 performs the $\aFinish$ action. This action checks that the expression inside the hole is compatible with the type the hole is being checked against, as indeed it now is, and removes the hole, resulting in the final edit state on Line 20. In practice, an editor might automatically perform the $\aFinish$ action as soon as it is possible to do so, but for simplicity, Hazelnut formally requires that it be performed explicitly.

>>>>

Figure~\ref{fig:first-example} shows an edit sequence that constructs the increment function, of type $\tarr{\tnum}{\tnum}$, starting from the empty hole via the indicated sequence of {actions}. We will introduce Hazelnut's formal syntax and define the referenced rules in Sec. \ref{sec:hazel}. First, let us build some high-level intuitions.

@@

The edit state in Hazelnut is a {Z-expression}, $\zexp$. Every Z-expression has a single {H-expression}, $\hexp$, or {H-type}, $\htau$, under the {cursor}, typeset $\zwsel{\hexp}$ or $\zwsel{\htau}$, respectively. For example, on Line 1, the empty expression hole, $\hhole{}$, is under the cursor.

@@

\begin{figure}[t!]
\begin{center}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
\hline
\# & \textbf{Z-Expression} & 
%\textbf{H-Expression} & 
% \textbf{Type} & 
\textbf{Next Action} & \textbf{Rule}
\\
\hline
1 &
\zwsel{\hhole{}} & 
% \hhole{} &
% \tehole 
% &
\aConstruct{\flam{x}} & 
\text{(\ref{r:conelamhole})}
\\ 2 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\hhole{}}}{\hhole{}} & 
% \hlam{x}{\hhole{}} : \tarr{\hhole{}}{\hhole{}} &
% \tarr{\tehole}{\tehole} &
\aConstruct{\fnum{}} &
\text{(\ref{r:contnum})}
\\ 3 &
\hlam{x}{\hhole{}} : \tarr{\zwsel{\tnum{}}}{\hhole{}} &
% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\hhole{}} &
% \tarr{\tnum}{\tehole} & 
\aMove{\dNext{}} & 
\text{({\ref{rule:move-nextSib-arr}})}
\\ 4 &
\hlam{x}{\hhole{}} : \tarr{\tnum}{\zwsel{\hhole{}}}
&
% \text{\textquotedbl}&
% \tarr{\tnum}{\tehole} & 
\aConstruct{\fnum{}} & 
\text{(\ref{r:contnum})}
\\ 5 &
\hlam{x}{\hhole{}} : \tarr{\tnum{}}{\zwsel{\tnum{}}} & 
% \hlam{x}{\hhole{}} : \tarr{\tnum{}}{\tnum{}} &
% \tarr{\tnum}{\tnum} &
\aMove{\dParent{}} & 
\text{(\ref{rule:move-parent-arr-right})}
\\ 6 &
\hlam{x}{\hhole{}} : \zwsel{\tarr{\tnum{}}{\tnum{}}}
&
% \text{\textquotedbl}&
% \tarr{\tnum}{\tnum} & 
\aMove{\dParent{}} & 
\text{(\ref{rule:move-parent-asc-right})}
\\ 7 & 
\zwsel{\hlam{x}{\hhole{}} : \tarr{\tnum}{\tnum}} & 
\aMove{\dChild} & 
\text{(\ref{r:movefirstchild})}
\\ 8 &
% &
\zwsel{\hlam{x}{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} & 
% \tarr{\tnum}{\tnum} &
% &
\aMove{\dChild{}} & 
\text{(\ref{r:movefirstchild-lam})}
\\ 9 &
% &
\hlam{x}{\zwsel{\hhole{}}} : \tarr{\tnum{}}{\tnum{}} &
% \tarr{\tnum}{\tnum} & 
\aConstruct{\fvar{x}} & 
\text{(\ref{r:conevar})}
\\ 10 &
% \hlam{x}{{x}} : \tarr{\tnum{}}{\tnum{}}
% &
\hlam{x}{\zwsel{{x}}} : \tarr{\tnum{}}{\tnum{}} &
% \tarr{\tnum}{\tnum} & 
{\aConstruct{\fplus}}
&
\text{(\ref{rule:construct-plus-compat})}
\\ 11 & 
\hlam{x}{\hadd{x}{\zwsel{\hhole{}}}} : \tarr{\tnum{}}{\tnum{}} & 
\aConstruct{\fnumlit{1}} & 
\text{(\ref{r:conenumnum})}
\\ 12 & 
\hlam{x}{\hadd{x}{\zwsel{\hnum{1}}}} : \tarr{\tnum}{\tnum} & 
\textrm{---} & 
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Constructing an increment function in Hazelnut.}
\label{fig:first-example}
\end{figure}

The actions on Lines 5-8 move the cursor to the function body. For simplicity, Hazelnut defines only a small collection of primitive movement actions. In practice, an editor would also define compound movement actions, e.g. an action that moves the cursor directly to the next remaining hole, in terms of these primitive movement actions.

Lines 9-11 complete the function body by first constructing the variable $x$, then  constructing the plus form, and finally constructing the number $\hnum{1}$. Notice that we did not need to construct the function body in an ``outside-in'' manner, i.e. we constructed $x$ before constructing the outer plus form. Luckily, the transient function bodies, $x$ and $x + \hhole{}$, could be checked against type $\tnum$ (as we will detail in Sec. \ref{sec:holes}.)

@@

\begin{figure}[t!]
\begin{center}
\colorbox{light-gray}{\hspace{53px} now assume $incr : \tarr{\tnum}{\tnum}$ \hspace{54px}}
% \vspace{-6px}
$\arraycolsep=4px
\begin{array}{|r||l|l||l|l|}
% \\
\hline
\# & \textbf{Z-Expression} & 
%\textbf{H-Expression} & 
% \textbf{Type} & 
\textbf{Next Action} & \textbf{Rule}
\\
\hline
13 &
\zwsel{\hhole{}} & 
% \hhole{} &
% \tehole 
% &
\aConstruct{\fvar{incr}} \hphantom{~\,\,~~}& 
\text{(\ref{r:conevar})}
\\ 14 & 
\zwsel{incr} & 
\aConstruct{\fap} & 
\text{(\ref{r:coneapfn})}
\\ 15 & 
incr(\zwsel{\hhole{}}) & 
\aConstruct{\fvar{incr}} & 
\text{(\ref{r:conevar2})}
\\ 16 & 
incr(\hhole{\zwsel{incr}}) & 
\aConstruct{\fap} & 
\text{(\ref{r:coneapfn})}
\\ 17 & 
incr(\hhole{incr(\zwsel{\hhole{}})}) \hphantom{~~~~} & 
\aConstruct{\fnumlit{3}} & 
\text{(\ref{r:conenumnum})}
\\ 18 & 
incr(\hhole{incr(\zwsel{\hnum{3}})}) & 
\aMove{\dParent}& 
\text{(\ref{r:moveparent-ap2})}
\\ 19 & 
incr(\hhole{\zwsel{incr(\hnum{3})}}) & 
\aMove{\dParent} & 
\text{(\ref{r:moveparent-hole})}
\\ 20 &
incr(\zwsel{\hhole{incr(\hnum{3})}})& 
\aFinish & 
\text{(\ref{r:finishana})}
\\ 21 & 
incr(\zwsel{incr(\hnum{3})}) & 
\textrm{---} & 
{\textrm{---}}
\\ \hline
\end{array}
$\end{center}\vspace{-6px}
\caption{Appyling the increment function.}
\label{fig:second-example}
\end{figure}

Figure \ref{fig:second-example} shows an edit sequence that constructs the expression $incr(incr(\hnum{3}))$, where $incr$ is assumed bound to the increment function from Figure \ref{fig:first-example}.

We begin on Line 13 by constructing the variable $incr$. Line 14 then constructs the application form with $incr$ in function position, leaving the cursor on a hole in the argument position. Notice that we did not need to construct the outer application form before identifying the function being applied. Again, the transient edit state happens to be well-typed, so we needed to make no special allowances for this order of actions.

We now need to apply $incr$ again, so we perform the same action on Line 15 as we did on Line 13, i.e. $\aConstruct{\fvar{incr}}$. In a syntactic structure editor, performing such an action would result in the following edit state: 
\[
incr(\zwsel{incr})
\]
This edit state is ill-typed (after \emph{cursor erasure}): the argument of $incr$ must be of type $\tnum$ but here it is of type $\tarr{\tnum}{\tnum}$. Hazelnut cannot allow such an edit state to arise.

The programmer could alternatively have performed the $\aConstruct{\fap}$ action on Line 15. This would result in the following edit state, which is well-typed according to the static semantics that we will define in the next section:
\[
incr(\hhole{}(\zwsel{\hhole{}}))
\]
The problem is that the programmer is not able to identify the intended function before constructing the outer function application form. This stands in contrast to Lines 13-14. % This is qualitatively unnatural.

Hazelnut's action semantics addresses this problem: rather than disallowing the $\aConstruct{\fvar{incr}}$ action on Line 15, it leaves $incr$ inside a hole:
\[
incr(\hhole{\zwsel{incr}})
\]
This defers the type consistency check, exactly as an empty hole in the same position does. One way to think about non-empty holes is as an internalization of the ``red underline'' that text or syntactic structure editors display to indicate a type inconsistency. By internalizing this concept, the presence of a type inconsistency does not leave the entire program formally meaningless.

The expression inside a non-empty hole must itself be well-typed, so the programmer can continue to edit it. Lines 16-17 proceed to apply the inner mention of $incr$ to a number literal, $\hnum{3}$. Finally, Lines 18-19 move the cursor to the non-empty hole and Line 20 performs the $\aFinish$ action. The $\aFinish$ action removes the hole if the type of the expression inside the hole is consistent with the expected type, as it now is. This results in the final edit state on Line 20, as desired. In practice, the editor might automatically perform the $\aFinish$ action as soon as it becomes possible, but for simplicity, Hazelnut formally requires that it be performed explicitly.
